---
title: "Random sampling from the Poisson kernel-based density"
author: "Giovanni Saraceno"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Random sampling from the Poisson kernel-based density}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r srr-tags, eval = FALSE, echo = FALSE}
#' srr tags
#' 
#' 
#' @srrstats {PD4.2} The results of the random sampling using the different 
#'                   distributions are compared graphically, by plotting on the 
#'                   sphere the generated observations.   
#'                   
```

```{r}
library(QuadratiK)
```

In this example, we generate observations from the Poisson kernel-based 
distribution on the sphere, $S^{d-1}$. 
We consider mean direction $\mu=(0,0,1)$, $d=3$ and the concentration
parameter is $\rho = 0.8$. 
```{r}
mu <- c(0,0,1)
d <- 3
n <- 1000
rho <- 0.8
```
We sampled $n=1000$ observations for each method available.

```{r}
n <- 1000
set.seed(2468)
# Generate observations using the rejection algorithm with von-Mises 
# distribution envelopes
dat1 <- rpkb(n = n, rho=rho, mu=mu, method="rejvmf")
# Generate observations using the rejection algorithm with angular central 
# Gaussian distribution envelopes
dat2 <- rpkb(n = n, rho=rho, mu=mu, method="rejacg")
# Generate observations using the projected Saw distribution
dat3 <- rpkb(n = n, rho=rho, mu=mu, method="rejpsaw")
```
The function \texttt{rpkb} returns a list with the following components 
```{r}
summary(dat1)
```
We extract the generated data sets and create a unique data set.
```{r}
x <- rbind(dat1$x, dat2$x, dat3$x)
```
Finally, the observations generated through the different methods are compared
graphically, by displaying the data points on the sphere colored with respect to
the used method.

```{r fig.width=6, fig.height=8}
library(rgl)
# Legend information
classes = c("rejvmf", "rejacg", "rejpsaw")
# Fix a color for each method
colors = c("red", "blue", "green")
labels <- factor(rep(colors, each = 1000))
# Element needed for the Legend position in the following plot
offset = 0.25
# Coordinates for legend placement
legend_x = max(x[,1]) + offset  
legend_y = max(x[,2]) + offset
legend_z = seq(min(x[,3]), length.out = length(classes), by = offset)

open3d()
# Create the legend
for (i in seq_along(classes)) {
   text3d(legend_x, legend_y, legend_z[i], texts = classes[i], adj = c(0, 0.5))
   points3d(legend_x-0.1, legend_y, legend_z[i], col = colors[i], size = 5)
}
title3d("", line = 3, cex = 1.5, font=2, add=TRUE)
# Plot the sampled observations colored with respect to the used method
plot3d(x[,1], x[,2], x[,3], col = labels, size = 5, add=TRUE)
title3d("", line = 3, cex = 1.5, font=2, add=TRUE)
# Add a Sphere as background
rgl.spheres(0 , col = "transparent", alpha = 0.2)
# Rotate and zoom the generated 3 dimensional plot to facilitate visualization
view3d(theta = 10, phi = -25, zoom = 0.5)
# rglwidget is added in order to display the generated figure into the html replication file.
rglwidget()
close3d()

```
