---
title: "Introduction to the QuadratiK Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the QuadratiK Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

The `QuadratiK` package provides the first implementation, in R and Python, of a comprehensive set of goodness-of-fit tests and a clustering technique for spherical data using kernel-based quadratic distances. The primary goal of `QuadratiK` is to offer flexible tools for testing multivariate and high-dimensional data for uniformity, normality, comparing two or more samples, and clustering spherical data efficiently.

This package includes several novel algorithms that are designed to handle spherical data, which is often encountered in fields like directional statistics, geospatial data analysis, and signal processing.

### Installation

You can install the version published on CRAN of `QuadratiK`:

```{r, eval = FALSE}
install.packages("QuadratiK")
```

Or the development version on GitHub:

```{r, eval = FALSE}
library(devtools)
devtools::install_github('giovsaraceno/QuadratiK-package')
```

The `QuadratiK` package is also available in Python on PyPI and as a Dashboard application. Usage instruction for the Dashboard can be found at <https://quadratik.readthedocs.io/en/latest/user_guide/dashboard_application_usage.html>.

### Citation

If you use this package in your research or work, please cite it as follows:

Saraceno G, Markatou M, Mukhopadhyay R, Golzy M (2024). QuadratiK: Collection of Methods Constructed using Kernel-Based Quadratic Distances. <https://cran.r-project.org/package=QuadratiK>

``` bibtex
@Manual{saraceno2024QuadratiK, 
   title = {QuadratiK: Collection of Methods Constructed using Kernel-Based 
            Quadratic Distances}, 
   author = {Giovanni Saraceno and Marianthi Markatou and Raktim Mukhopadhyay 
            and Mojgan Golzy}, 
   year = {2024}, 
   note = {<https://cran.r-project.org/package=QuadratiK>,
   <https://github.com/giovsaraceno/QuadratiK-package>, 
   <https://giovsaraceno.github.io/QuadratiK-package/>} 
}
```

and the associated paper:

Saraceno Giovanni, Markatou Marianthi, Mukhopadhyay Raktim, Golzy Mojgan (2024). Goodness-of-Fit and Clustering of Spherical Data: the QuadratiK package in R and Python. arXiv preprint arXiv:2402.02290.

``` bibtex
@misc{saraceno2024package, 
   title={Goodness-of-Fit and Clustering of Spherical Data: the QuadratiK package in
          R and Python}, 
   author={Giovanni Saraceno and Marianthi Markatou and Raktim Mukhopadhyay and
           Mojgan Golzy}, 
   year={2024}, 
   eprint={2402.02290}, 
   archivePrefix={arXiv}, 
   primaryClass={stat.CO},    
   url={<https://arxiv.org/abs/2402.02290>}
}
```

## Key features and basic usage

### Goodness-of-Fit Tests

```{r}
library(QuadratiK)
```

The software implements one, two, and *k*-sample tests for goodness of fit, offering an efficient and mathematically sound way to assess the fit of probability distributions. Our tests are particularly useful for large, high dimensional data sets where the assessment of fit of probability models is of interest.

The provided goodness-of-fit tests can be performed using the `kb.test()` function. The kernel-based quadratic distance tests are constructed using the normal kernel which depends on the tuning parameter $h$. If a value for $h$ is not provided, the function perform the `select_h()` algorithm searching for an optimal value. For more details please visit the relative help documentations.

```{r, eval=FALSE}
?kb.test
?select_h
```

The proposed tests perform well in terms of level and power for contiguous alternatives, heavy tailed distributions and in higher dimensions.

**Test for normality**

To test the null hypothesis of normality $H_0:F=\mathcal{N}_d(\mu, \Sigma)$

```{r}
x <- matrix(rnorm(100), ncol = 2)
# Does x come from a multivariate standard normal distribution?
kb.test(x, h=0.4)
```

If needed, we can specify $\mu$ and $\Sigma$, otherwise the standard normal distribution is considered.

```{r}
x <- matrix(rnorm(100,4), ncol = 2)
# Does x come from a multivariate standard normal distribution?
kb.test(x, mu_hat = c(4,4), Sigma_hat = diag(2), h = 0.4)
```

**Two-sample test**

In case we want to compare two samples $X \sim F$ and $Y \sim G$ with the null hypothesis $H_0:F=G$ vs $H_1:F\not =G$.

```{r}
x <- matrix(rnorm(100), ncol = 2)
y <- matrix(rnorm(100,mean = 5), ncol = 2)
# Do x and y come from the same distribution?
kb.test(x, y, h = 0.4)
```

**k-sample test**

In case we want to compare $k$ samples, with $k>2$, that is $H_0:F_1=F_2=\ldots=F_k$ vs $H_1:F_i\not =F_j$ for some $i\not = j$.

```{r}
x1 <- matrix(rnorm(100), ncol = 2)
x2 <- matrix(rnorm(100), ncol = 2)
x3 <- matrix(rnorm(100, mean = 5), ncol = 2)
y <- rep(c(1, 2, 3), each = 50)
# Do x1, x2 and x3 come from the same distribution?
x <- rbind(x1, x2, x3)
kb.test(x, y, h = 0.4)
```

#### Test for uniformity on the sphere

Expanded capabilities include supporting tests for uniformity on the *d*-dimensional Sphere based on Poisson kernel. The Poisson kernel depends on the concentration parameter $\rho$ and a location vector $\mu$. For more details please visit the help documentation of the `pk.test()` function.

```{r, eval=FALSE}
?pk.test
```

To test the null hypothesis of uniformity on the $d$-dimensional sphere $\mathcal{S}^{d-1} = \{x \in \mathbb{R}^d : ||x||=1 \}$

```{r}
# Generate points on the sphere from the uniform ditribution 
x <- sample_hypersphere(d = 3, n_points = 100)
# Does x come from the uniform distribution on the sphere?
pk.test(x, rho = 0.9)
```

### Poisson kernel-based distribution (PKBD)

The package offers functions for computing the density value and for generating random samples from a PKBD. The Poisson kernel-based densities are based on the normalized Poisson kernel and are defined on the $d$-dimensional unit sphere. For more details please visit the help documentation of the `dpkb()` and `rpkb()` functions.

```{r, eval=FALSE}
?dpkb
?rpkb
```

*Example*

```{r}
mu <- c(1,0,0)
rho <- 0.9
x <- rpkb(n = 100, mu = mu, rho = rho)
head(x$x)
dens_x <- dpkb(x$x, mu = mu, rho = rho)
head(dens_x)
```

### Clustering Algorithm for Spherical Data

The package incorporates a unique clustering algorithm specifically tailored for spherical data and it is especially useful in the presence of noise in the data and the presence of non-negligible overlap between clusters. This algorithm leverages a mixture of Poisson kernel-based densities on the Sphere, enabling effective clustering of spherical data or data that has been spherically transformed. For more details please visit the help documentation of the `pkbc()` function.

```{r, eval=FALSE}
?pkbc
```

*Example*

```{r}
# Generate 3 samples from the PKBD with different location directions
x1 <- rpkb(n = 100, mu = c(1,0,0), rho = rho)
x2 <- rpkb(n = 100, mu = c(-1,0,0), rho = rho)
x3 <- rpkb(n = 100, mu = c(0,0,1), rho = rho)
x <- rbind(x1$x, x2$x, x3$x)
# Perform the clustering algorithm
# Serch for 2, 3 or 4 clusters
cluster_res <- pkbc(dat = x, nClust = c(2, 3, 4))
summary(cluster_res)
```

The software includes additional graphical functions, aiding users in validating and representing the cluster results as well as enhancing the interpretability and usability of the analysis.

```{r}
# Predict the membership of new data with respect to the clustering results
x_new <- rpkb(n = 10, mu = c(1,0,0), rho = rho)
memb_mew <- predict(cluster_res, k = 3, newdata = x_new$x)
memb_mew$Memb

```

```{r}
# Compute measures for evaluating the clustering results
val_res <- pkbc_validation(cluster_res)
val_res
```

```{r, fig.show = "hold", fig.width=6, fig.height=6}
# Plot method for the pkbc object:
# - scatter plot of data points on the sphere
# - elbow plot for helping the choice of the number of clusters
plot(cluster_res)
```

## Additional Resources

For more detailed information about the `QuadratiK` package, you can explore the following resources:

-   [Package Documentation on CRAN](https://cloud.r-project.org/web/packages/QuadratiK/index.html) – Official package documentation on CRAN.
-   [GitHub Repository](https://github.com/giovsaraceno/QuadratiK-package) – The GitHub repository with the development version, issues, and community discussions.
-   [QuadratiK Package Website](https://giovsaraceno.github.io/QuadratiK-package/) – A dedicated website with additional tutorials and examples.

If you're new to the package, we recommend starting with the available vignettes:

-   [Two-sample test](https://giovsaraceno.github.io/QuadratiK-package/articles/TwoSample_test.html)
-   [k-sample test](https://giovsaraceno.github.io/QuadratiK-package/articles/kSample_test.html)
-   [Test for uniformity](https://giovsaraceno.github.io/QuadratiK-package/articles/uniformity.html)
-   [Clustering on the sphere](https://giovsaraceno.github.io/QuadratiK-package/articles/wireless_clustering.html)
-   [Generate from PKBD](https://giovsaraceno.github.io/QuadratiK-package/articles/generate_rpkb.html)

### References

For more information on the methods implemented in this package, refer to the associated research papers:

-   Markatou Marianthi & Saraceno Giovanni (2024). “A Unified Framework for Multivariate Two- and k-Sample Kernel-based Quadratic Distance Goodness-of-Fit Tests.” [arXiv:2407.16374](https://doi.org/10.48550/arXiv.2407.16374)

-   Ding Yuxin, Markatou Marianthi, Saraceno Giovanni (2023). “Poisson Kernel-Based Tests for Uniformity on the d-Dimensional Sphere.” Statistica Sinica. doi: [10.5705/ss.202022.0347](https://doi.org/10.5705/ss.202022.0347).

-   Mojgan Golzy & Marianthi Markatou (2020) Poisson Kernel-Based Clustering on the Sphere: Convergence Properties, Identifiability, and a Method of Sampling, Journal of Computational and Graphical Statistics, 29:4, 758-770, DOI: [10.1080/10618600.2020.1740713](https://doi.org/10.1080/10618600.2020.1740713).
