---
title: "k-sample test"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{k-sample test}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
```{r srr-tags, eval = FALSE, echo = FALSE}
#' srr tags
#' 
#' 
#' @srrstats {G1.5} k-sample test example in the associated paper
```            


Consider $k$ random samples of i.i.d. observations 
$\mathbf{x}^{(i)}_1, \mathbf{x}^{(i)}_{2}, \ldots, \mathbf{x}^{(i)}_{n_i} \sim F_i$, for $i = 1, \ldots, k$. 

We test if the samples are generated from the same *unknown* distribution $\bar{F}$, that is:
$$
H_0: F_1 = F_2 = \ldots = F_k = \bar{F}
$$
versus the alternative where two of the $k$ distributions differ, that is
$$
H_1: F_i \not = F_j,
$$
for some $1 \le i \not = j \le k$.

Upon the construction of a matrix distance $\hat{\mathbf{D}}$, with off-diagonal elements 
$$
	\hat{D}_{ij} = \frac{1}{n_i n_j} \sum_{\ell=1}^{n_i}\sum_{r=1}^{n_j}K_{\bar{F}}(\mathbf{x}^{(i)}_\ell,\mathbf{x}^{(j)}_r), \qquad \mbox{ for }i \not= j
$$
and in the diagonal
$$	
\hat{D}_{ii} = \frac{1}{n_i (n_i -1)} \sum_{\ell=1}^{n_i}\sum_{r\not= \ell}^{n_i}K_{\bar{F}}(\mathbf{x}^{(i)}_\ell,\mathbf{x}^{(i)}_r), \qquad \mbox{ for }i = j,
$$
where $K_{\bar{F}}$ denotes the Normal kernel $K$, defined as

$$
K(\mathbf{s}, \mathbf{t}) = (2 \pi)^{-d/2} 
\left(\det{\mathbf{\Sigma}_h}\right)^{-\frac{1}{2}}  
\exp\left\{-\frac{1}{2}(\mathbf{s} - \mathbf{t})^\top 
\mathbf{\Sigma}_h^{-1}(\mathbf{s} - \mathbf{t})\right\},
$$

for every $\mathbf{s}, \mathbf{t} \in \mathbb{R}^d \times 
\mathbb{R}^d$, with covariance matrix $\mathbf{\Sigma}_h = h^2 I$ and
tuning parameter $h$, centered with respect to 
$$
\bar{F} = \frac{n_1 \hat{F}_1 + \ldots + n_k \hat{F}_k}{n}, \quad \mbox{ with } n=\sum_{i=1}^k n_i.
$$
For more information about the centering of the kernel, see the documentation of the ``kb.test()`` function.
```{r eval=FALSE} 
help(kb.test) 
```
We compute the trace statistic
$$	
\mathrm{trace}(\hat{\mathbf{D}}_n) =  \sum_{i=1}^{k}\hat{D}_{ii}.
$$
and $D_n$, derived considering all the possible pairwise comparisons in the $k$-sample null hypothesis, given as
$$	
D_{n} = (k-1) \mathrm{trace}(\hat{\mathbf{D}}_n) - 2 \sum_{i=1}^{k}\sum_{j> i}^{k}\hat{D}_{ij}.
$$

We show the usage of the ``kb.test()`` function with the following example of $k=3$ samples of bivariate observations following normal distributions with different mean vectors. 

We generate three samples, with $n=50$ observations each, from a 2-dimensional Gaussian distributions with mean vectors $\mu_1 = (0, \frac{\sqrt{3}}{3})$, ${\mu}_2 = (-\frac{1}{2}, -\frac{\sqrt{3}}{6})$ and $\mu_3 = (\frac{1}{2}, -\frac{\sqrt{3}}{6})$, and the Identity matrix as covariance matrix. In this situation, the generated samples are well separated, following different Gaussian distributions, i.e. $X_1 \sim N_2(\mu_1, I)$, $X_2 \sim N_2(\mu_2, I)$ and $X_3 \sim N_2(\mu_3, I)$}. 
In order to perform the $k$-sample tests, we need to define the vector ``y`` which  indicates the membership to groups.  
```{r, message=FALSE, warning=FALSE}
library(mvtnorm)
library(QuadratiK)
library(ggplot2)
sizes <- rep(50,3)
eps <- 1
set.seed(2468)
x1 <- rmvnorm(sizes[1], mean = c(0,sqrt(3)*eps/3))
x2 <- rmvnorm(sizes[2], mean = c(-eps/2,-sqrt(3)*eps/6))
x3 <- rmvnorm(sizes[3], mean = c(eps/2,-sqrt(3)*eps/6))
x <- rbind(x1, x2, x3)
y <- as.factor(rep(c(1,2,3), times=sizes))
```
```{r fig.width=6, fig.height=4}
ggplot(data.frame(x=x, y=y), aes(x = x[,1], y = x[,2], color = y)) +
  geom_point(size = 2) +
  labs(title = "Generated Points", x = "X1", y = "X2") +
  theme_minimal()
```

To use the ``kb.test()`` function, we need to provide the value for the tuning parameter $h$. The function ``select_h`` can be used for identifying on *optimal* value of $h$. This function needs the input ``x`` and ``y`` as the function ``kb.test``, and the selection of the family of alternatives. Here we consider the location alternatives. 
```{r fig.width=6, fig.height=4}
set.seed(2468)
h_k <- select_h(x=x, y=y, alternative="location")
```
```{r}
h_k$h_sel
```
The ``select_h`` function has also generated a figure displaying the obtained power versus the considered $h$, for each value of alternative $\delta$ considered.

We can now perform the $k$-sample tests with the *optimal* value of $h$. 
```{r}
set.seed(2468)
k_test <- kb.test(x=x, y=y, h=h_k$h_sel)
show(k_test)
```
The function ``kb.test()`` returns an object of class ``kb.test``.
The ``show`` method for the ``kb.test`` object shows the computed statistics with corresponding critical values, and the logical indicating if the null hypothesis is rejected. 
The test correctly rejects the null hypothesis, in fact the values of the statistics are greater than the computed critical values. 
The package provides also the ``summary`` function which returns the results of the tests together with the standard descriptive statistics for each variable computed, overall, and with respect to the provided groups. 
```{r}
summary_ktest <- summary(k_test)
summary_ktest$summary_tables
```

#### Note

If a value of $h$ is not provided to ``kb.test()``, this function performs the function ``select_h`` for automatic search of an *optimal* value of $h$ to use. . The following code shows its usage, but it is not executed since we would obtain the same results.
```{r eval=FALSE}
k_test_h <- kb.test(x=x, y=y)
```
For more details visit the help documentation of the ``select_h()`` function.
```{r eval=FALSE} 
help(select_h) 
```

In the ``kb.test()`` function, the critical value can be computed with the subsampling, bootstrap or permutation algorithm. The default method is set to subsampling since it needs less computational time. For details on the sampling algorithm see the documentation of the ``kb.test()`` function and the following reference.

The proposed tests exhibit high power against asymmetric alternatives that are close to the null hypothesis and with small sample size, as well as in the $k \ge 3$ sample comparison, for dimension $d>2$ and all sample sizes. For more details, see the extensive simulation study reported in the following reference.

## References

Markatou, M. and Saraceno, G. (2024). “A Unified Framework for 
Multivariate Two- and k-Sample Kernel-based Quadratic Distance 
Goodness-of-Fit Tests.”  
[https://doi.org/10.48550/arXiv.2407.16374](https://doi.org/10.48550/arXiv.2407.16374)
