---
title: "Uniformity test on the Sphere"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Uniformity test on the Sphere}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

### Uniformity test on the Sphere
```{r srr-tags, eval = FALSE, echo = FALSE}
#' srr tags
#' 
#' 
#' @srrstats {G1.5} Uniformity test example in the associated paper
```
The quadratic distance between two probability distributions $F$ and $G$ is
then defined as:

$$
d_{K}(F,G) = \int\int K(x,y) d(F-G)(x) d(F-G)(y),
$$

where $G$ is a distribution whose goodness of fit we wish to assess, that
is the uniform distribution on the sphere $\mathcal{S}^{d-1}$, and 
$K$ denotes the Poisson kernel defined as:

$$
K(\mathbf{u}, \mathbf{v}) = \frac{1-\rho^{2}}{\left(1+\rho^{2}-2\rho
(\mathbf{u}\cdot \mathbf{v})\right)^{d/2}},
$$

for every $\mathbf{u}, \mathbf{v} \in \mathcal{S}^{d-1} \times 
\mathcal{S}^{d-1}$.

Let $x_1, x_2, \ldots, x_n$ be a random sample with empirical distribution
function $\hat{F}$.

We test the null hypothesis of uniformity on the 
$d$-dimensional sphere, i.e., $H_0: F = G$, by considering the U-statistic 
estimate and the V-statistic estimate of the sample KBQD (Kernel-Based Quadratic Distance).

We generated $n=200$ observations from the uniform distribution on $S^{d-1}$, with $d=3$, and $\rho = 0.7$. 
```{r, message=FALSE}
library(QuadratiK)
n <- 200
d <- 3
set.seed(2468)
z <- matrix(rnorm(n * d), n, d)
dat_sphere <- z/sqrt(rowSums(z^2))
```

The ``pk.test`` is used for testing uniformity of the generated sample, providing the data matrix as ``x`` and the value of the concentration parameter ``rho``. 
```{r}
rho <- 0.7
set.seed(2468)
res_unif <- pk.test(x=dat_sphere, rho=rho)

show(res_unif)
```
The ``pk.test()`` function returns an object of class ``pk.test``. The ``show`` method displays the computed statistics and the corresponding critical values. 

The test correctly does not reject the null hypothesis of uniformity.

The ``summary`` method for the ``pk.test`` output object provides the results of the performed test, and generates a figure showing the qq-plots against the uniform distribution of each variable with a table of standard descriptive statistics. 
```{r fig.width=6, fig.height=8}
summary_unif <- summary(res_unif)
```

The figure automatically generated by the ``summary`` function on the result of the test for uniformity displays the  qq-plots between the given samples and the uniform distribution with a table of the standard descriptive statistics for each variable.

#### Multimodal example

The Poisson kernel-based test for uniformity exhibits excellent results especially in the case of multimodal distributions. 

In this example, we generate data points from a mixture of 4 von Mises-Fisher distributions in 2 dimensions. The direction of the mean vectors for the distributions are set as (1, 0), (0, 1), (-1, 0), and (0, -1), with a concentration parameter $\kappa=5$. 

```{r, warning=FALSE, message=FALSE}
# Load necessary libraries
library(movMF)        
library(sphunif)      
```

```{r}
set.seed(2468)
# Define the mean directions of the 4 von Mises-Fisher distributions
means <- rbind(
  c(1, 0),    
  c(0, 1),    
  c(-1,0),   
  c(0, -1)    
)
# Define the concentration parameter (kappa)
kappa <- 5
# Generate 100 samples from a mixture of 4 von Mises-Fisher distributions
samples <- matrix(rmovMF(100, theta = kappa * means), ncol=2)
```

We now compare the results of the `pk.test` function with the Ajne and Bingham tests using the `sphunif` package.

```{r}
# Run the pk.test from the QuadratiK package to test the data
pk_test_result <- pk.test(samples, rho=0.8)

# Run the Bingham and Ajne tests from the sphunif package
other_test_result <- unif_test(samples, type = c("Bingham", "Ajne"))

pk_test_result
other_test_result

```
In this example, the Poisson kernel-based test statistics reject the null hypothesis, while the Bingham test and Ajne test obtain p-values equal to 0.0532 and 0.3156, respectively. 

## References

Ding, Y., Markatou, M., Saraceno, G. (2023). "Poisson Kernel-Based Tests for
Uniformity on the d-Dimensional Sphere." *Statistica Sinica*. 
[doi:10.5705/ss.202022.0347](https://doi.org/10.5705/ss.202022.0347).
